Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-19 paper.tex

\begin{abstract}
Recombination is an important and ongoing feature of SARS-CoV-2 evolution,
and presents major challenges to [things important to public health].
Detection of new recombinant strains is challenging, with methods
[being a bit rubbish].
In addition, many basic approaches fundamental to the study of viral
evolution assume that recombination is negligible, in that a single
phylogenetic tree can represent the genetic ancestry of the
circulating strains. Here we present an initial version of
\texttt{sc2ts}, a method to automatically detect recombinants
in real-time and to fully integrate them into a cohesive joint
model of mutation, recombination and genetic inheritance known
as an Ancestral Recombination Graph (ARG). We infer two ARGs under
different sampling strategies, and study their properties.
One is densely sampled, and includes 1.27 million sequences
sampled up to June 30, 2021.
The second is more sparsely sampled,
consisting of 784K sequences sampled up to June 30, 2022.
We find that both ARGs are
% better words?
highly consistent with
known features of SARS-CoV-2 evolution, recovering the basic
backbone phylogeny, mutational spectra, and recapitulating
details on the majority of known recombinant lineages.
Using the well-established and fully-featured \texttt{tskit} library,
the ARGs can also be stored concisely and processed efficiently
using standard Python tools. For example, the ARG for 1.27 million
sequences---encoding the inferred reticulate ancestry,
full alignments, and extensive metadata---requires
58MB of storage,
and loads in less than a second.
%JK Not sure if we'd actually say it like this, but this is the
% basic gist I'd like to get accross[
The ability to fully integrate the effects of recombination into
downstream analyses, to quickly and automatically detect new recombinants,
and to provide an efficient and convenient platform for computation
makes \texttt{sc2ts} a promising approach, worthy of
further investment and development.
\end{abstract}

>>>>

\begin{abstract}
Recombination is an important and ongoing feature of SARS-CoV-2 evolution,
and presents major challenges to [things important to public health].
Detection of new recombinant strains is challenging, with methods
[being a bit rubbish].
In addition, many basic approaches fundamental to the study of viral
evolution assume that recombination is negligible, in that a single
phylogenetic tree can represent the genetic ancestry of the
circulating strains. Here we present an initial version of
\texttt{sc2ts}, a method to automatically detect recombinants
in real-time and to fully integrate them into a cohesive joint
model of mutation, recombination and genetic inheritance known
as an Ancestral Recombination Graph (ARG). We infer two ARGs under
different sampling strategies, and study their properties.
One is densely sampled, and includes 1.27 million sequences
sampled up to June 30, 2021.
The second is more sparsely sampled,
consisting of 784K sequences sampled up to June 30, 2022.
We find that both ARGs are
% better words?
highly consistent with
known features of SARS-CoV-2 evolution, recovering the basic
backbone phylogeny, mutational spectra, and recapitulating
details on the majority of known recombinant lineages.
Using the well-established and featured-rich \texttt{tskit} library,
the ARGs can also be stored concisely and processed efficiently
using standard Python tools. For example, the ARG for 1.27 million
sequences---encoding the inferred reticulate ancestry,
full alignments, and extensive metadata---requires
58MB of storage,
and loads in less than a second.
%JK Not sure if we'd actually say it like this, but this is the
% basic gist I'd like to get accross[
The ability to fully integrate the effects of recombination into
downstream analyses, to quickly and automatically detect new recombinants,
and to provide an efficient and convenient platform for computation
makes \texttt{sc2ts} a promising approach, worthy of
further investment and development.
\end{abstract}
